---
title: "Constructing a time series using MODIS data"
author: "John Mutua"
date: "Monday, September 05, 2016"
use: "Script extracts NDVI values from MODIS rasters and plots a time series by season and month"
data input: "Preprocessd MODIS rasters"
output: html_document

---

First, clear your workspace

```{r}
rm(list = ls(all = TRUE))
```

Second, load all required libraries
```{r}
require(raster) #working with rasters#
require(rgdal) #working with geospatial data#
require(ggplot2) #creating graphs#
require(dplyr) #subsetting by season#
require(lubridate) #working with dates#
require(grid) #arranging plots#
require(gridExtra) #arranging plots#
```

Set your working directory
```{r}
setwd("D:/ToBackup/Data/Regional/LDN_Namibia")
```

Import-NDVI-rasters
```{r}
rasList <- list.files("NDVI_Modis/AOI_NDVI", full.names = TRUE, pattern = ".tif$")
```

Create a raster stack
```{r}
NDVI_Otjo_stack <- stack(rasList) 
```

Insert a Temporal id to the dataframe. Characters corresponding to the year are located between the 20 and 23 position
while characters corresponding to the date are located between the 24 and 26 position
```{r}
oldnames<-names(NDVI_Otjo_stack)
head(oldnames)
year<-substr(oldnames,20,23)
table(year)
julianDay<-substr(oldnames,24,26)
```

Calculate mean NDVI for each raster and convert output array to a data frame
```{r}
avg_NDVI_Otjo <- as.data.frame(cellStats(NDVI_Otjo_stack,mean))
```

Rename the NDVI column
```{r}
names(avg_NDVI_Otjo) <- "meanNDVI"
```

Add julianDay values as a column in the data frame
```{r}
avg_NDVI_Otjo$julianDay <- julianDay
```

Check out the class in the new column
```{r}
class(avg_NDVI_Otjo$julianDay)
```

Create a time vector with the date
```{r}
tVector<-paste(year,julianDay,sep="-")
```

# convert the vector to a date
timeNDVI<-as.Date(tVector,format = "%Y-%j")
head(timeNDVI)

# add date values as a column in the data frame
avg_NDVI_Otjo$date <- timeNDVI

# Classes of the two columns now? 
class(avg_NDVI_Otjo$date)
class(avg_NDVI_Otjo$julianDay)

# add a site column to our data
avg_NDVI_Otjo$site <- "Otjozondjupa Region"

# view data
head(avg_NDVI_Otjo)

# add a "year" column to our data
avg_NDVI_Otjo$year <- year

# view data
avg_NDVI_Otjo

## ----plot-time-series----
# plot NDVI by time although this doesn't make sense

ggplot(avg_NDVI_Otjo, aes(year, meanNDVI), na.rm=TRUE) +
  geom_point(size=4,colour = "PeachPuff4") + 
  ggtitle("MODIS Derived NDVI - Otjozondjupa Region") +
  xlab("Time") + ylab("Mean NDVI") +
  theme(text = element_text(size=20))

## write NDVI data to a .csv File
# create new df to prevent changes to avg_NDVI_Otjo
Otjo_NDVI_Values<-avg_NDVI_Otjo

# drop the row.names column 
row.names(Otjo_NDVI_Values)<-NULL

# check data frame
head(Otjo_NDVI_Values)

# create a .csv of mean NDVI values 
write.csv(Otjo_NDVI_Values, file="meanNDVI_Otjo_2005-2015.csv")

# add month to data frame
avg_NDVI_Otjo$month  <- month(avg_NDVI_Otjo$date)

# view head and tail of column
head(avg_NDVI_Otjo$month)
tail(avg_NDVI_Otjo$month)

# subset by data by Season
# create a new categorical variable called season by grouping months together.
avg_NDVI_Otjo_Seasons <- avg_NDVI_Otjo %>% 
  mutate(season = 
           ifelse(month %in% c(12, 1, 2, 3, 4, 5), "Hot-Wet",
                  ifelse(month %in% c(6, 7, 8), "Cool-Dry",
                         ifelse(month %in% c(9, 10, 11), "Hot-Dry", "Error"))))

# check to see if this worked
head(avg_NDVI_Otjo_Seasons$month)
head(avg_NDVI_Otjo_Seasons$season)
tail(avg_NDVI_Otjo_Seasons$month)
tail(avg_NDVI_Otjo_Seasons$season)

######################################
###-----------MONTHLY PLOTS--------###
######################################

# aggregate data by month
monthNDVI<-avg_NDVI_Otjo_Seasons %>%
  group_by(month) %>%
  summarise(monthNDVI=mean(meanNDVI, na.rm=TRUE))

# convert month numeric to month abbrevaition
monthNDVI$month_name <- month.abb[monthNDVI$month]

# reassign the month_name field to a factor
monthNDVI$month_name = factor(monthNDVI$month_name,
                              levels=c('Jan','Feb','Mar',
                                       'Apr','May','Jun','Jul',
                                       'Aug','Sep','Oct',
                                       'Nov','Dec'))

head(monthNDVI)

# plot data by month
Monthly_NDVI_Plot<-ggplot(monthNDVI, aes(month_name, monthNDVI, group=4)) +
  geom_line(colour="red") +
  ggtitle("Average NDVI (2005-2015) - Monthly") +
  xlab("Month") + ylab("Mean NDVI") +
  theme(plot.title = element_text(lineheight=.8, face="bold",
                                  size = 20)) +
  theme(text = element_text(size=18)) + geom_point()
Monthly_NDVI_Plot

# plot the same plot as before but with one plot per season, save as .pdf
ggsave(file="Otjozondjupa_Monthly_NDVI.pdf", width = 297, height = 210, units = "mm")

###----------INDIVINDUAL SEASONAL PLOTS------------###
### HOT-WET Season
Hot_Wet <- subset(monthNDVI, month >= 12 | month <= 5)
target <- c("12", "5", "4", "3", "2", "1")
Hot_Wet<-Hot_Wet[match(target, Hot_Wet$month),]

Hot_Wet$month_name <- factor(Hot_Wet$month_name, c("Dec", "Jan", "Feb", "Mar", "Apr", "May"))

Hot_Wet_NDVI_Plot<-ggplot(Hot_Wet, aes(month_name, monthNDVI, group=1)) +
  geom_line(colour="red") +
  ggtitle("Average NDVI (2005-2015) - Hot Wet Season ") +
  xlab("Month") + ylab("Mean NDVI") +
  theme(plot.title = element_text(lineheight=.8, face="bold",
                                  size = 20)) +
  theme(text = element_text(size=18)) + geom_point()
Hot_Wet_NDVI_Plot

# plot the same plot as before but with one plot per season, save as .pdf
ggsave(file="Otjozondjupa_Hot_Wet_NDVI.pdf", width = 297, height = 210, units = "mm")

### COOL-DRY Season
Cool_Dry <- subset(monthNDVI, month <= 8 & month >= 6)

Cool_Dry_NDVI_Plot<-ggplot(Cool_Dry, aes(month_name, monthNDVI, group=1)) +
  geom_line(colour="red") +
  ggtitle("Average NDVI (2005-2015) - Cool Dry Season") +
  xlab("Month") + ylab("Mean NDVI") +
  theme(plot.title = element_text(lineheight=.8, face="bold",
                                  size = 20)) +
  theme(text = element_text(size=18)) + geom_point()
Cool_Dry_NDVI_Plot

# plot the same plot as before but with one plot per season, save as .pdf
ggsave(file="Otjozondjupa_Cool_Dry_NDVI.pdf", width = 297, height = 210, units = "mm")

### HOT DRY Season
Hot_Dry <- subset(monthNDVI, month <= 11 & month >= 9)

Hot_Dry_NDVI_Plot<-ggplot(Hot_Dry, aes(month_name, monthNDVI, group=1)) +
  geom_line(colour="red") +
  ggtitle("Average NDVI (2005-2015) - Hot Dry Season") +
  xlab("Month") + ylab("Mean NDVI") +
  theme(plot.title = element_text(lineheight=.8, face="bold",
                                  size = 20)) +
  theme(text = element_text(size=18)) + geom_point()
Hot_Dry_NDVI_Plot

# plot the same plot as before but with one plot per season, save as .pdf
ggsave(file="Otjozondjupa_Hot_Dry_NDVI.pdf", width = 297, height = 210, units = "mm")

##-----------END OF MONTHLY PLOTS--------

######################################
##-----------SEASONAL PLOTS--------
######################################

# aggregate data by season
seasonNDVI<-avg_NDVI_Otjo_Seasons %>%
  group_by(year, season) %>%
  summarise(seasonNDVI=mean(meanNDVI, na.rm=TRUE))

# create a .csv of seasonal NDVI values 
write.csv(seasonNDVI, file="Seasonal_NDVI_2005-2015.csv")

# plot data by season
Seasonal_NDVI_Plot<-ggplot(seasonNDVI, aes(year, seasonNDVI, group=3)) +
  geom_line(colour="red") +
  ggtitle("Average NDVI (2005-2015) - Seasonal") +
  xlab("Year") + ylab("Mean NDVI") +
  theme(plot.title = element_text(lineheight=.8, face="bold",
                                  size = 20)) +
  theme(text = element_text(size=18)) + geom_point()

# plot the same plot as before but with one plot per season, save as .pdf
Seasonal_NDVI_Plot + facet_grid(. ~ season)
ggsave(file="Otjozondjupa_Seasonal_NDVI.pdf", width = 297, height = 210, units = "mm")

# plot the same plot in a landscape orientation, save as .pdf
Seasonal_NDVI_Plot + facet_grid(season ~ .)
ggsave(file="Otjozondjupa_Seasonal_NDVI.pdf", width = 297, height = 210, units = "mm")
